### MCQ

```{r quiz}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```

### MCQ with custom messages
```{r letter-a, echo=FALSE}
question("What number is the letter A in the *English* alphabet?",
  answer("8"),
  answer("1", correct = TRUE),
  answer("2", message = "2 is close but it's the letter B rather than A."),
  answer("26")
)
```


### MCQ extras
```{r letter-a-extras, echo=FALSE}
question("What number is the letter A in the English alphabet?",
  answer("8"),
  answer("14"),
  answer("1", correct = TRUE),
  answer("23"),
  random_answer_order = TRUE,
  allow_retry = TRUE
)
```

### Exercise with Code

*Here's an exercise with some prepopulated code as well as `exercise.lines = 5` to provide a bit more initial room to work.*

Now write a function that adds any two numbers and then call it:

```{r add-function, exercise=TRUE, exercise.lines = 5}
add <- function() {
  
}

add <- function(x,y){
  return(x+y)
}
add(1,2)
```

### Exercise with Hint

*Here's an exercise where the chunk is pre-evaluated via the `exercise.eval` option (so the user can see the default output we'd like them to customize). We also add a "hint" to the correct solution via the chunk immediate below labeled `print-limit-hint`.*

Modify the following code to limit the number of rows printed to 5:

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
mtcars
```

```{r print-limit-hint}
head(mtcars)
```