---
title: "Introduction to R for applied epidemiology"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Basics of R Syntax
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(error = TRUE)                    # set global options, can be overwritten by chunks
pacman::p_load(learnr, here, rio, tidyverse)           # load packages
linelist <- rio::import(here::here("www", "linelist_cleaned.rds")) # load data
tutorial_options(exercise.eval = FALSE)                 # pre-evaluate exercises
```


## Welcome to Intro to R for Applied Epidemiology
 
In this interactive tutorial, you will learn the basics of R in the context of tasks common to applied epidemiology:  

* Becoming familiar with typing code "commands"
* "Importing" a dataset into R to conduct cleaning and other modifications
* Viewing a dataset in R
* Reviewing and accessing columns in a dataset
* Key operators and functions
* Piping and pipe operators
* Errors and warnings



## Install and Load R packages

To get in some practice of basic R functions, we'll use the `linelist` fictional ebola outbreak data set you met in the EpiRHandbook.

To use basic functions with public health data, the tidyverse metapackage is very useful. Tidyverse loads the dplyr, ggplot2, and other packages that are useful in epi data analysis. 

We've preloaded the below packages for now. Installation and loading of these packages is described on the EpiRHandbook [_Suggested Packages_](https://epirhandbook.com/en/suggested-packages-1.html) page. 

```{r eval= FALSE}
pacman::p_load(learnr, here, rio, tidyverse)           # install and load R packages
```


### Import from subfolder

*Now it's your turn. Here's a simple exercise with an empty code chunk provided for entering the answer.*

Write the R code required to read in the original data object `linelist_cleaned.rds` and call it `linelist`. Then view the first few rows of the new object.

```{r exercise2, exercise=TRUE}

```

```{r exercise2-solution}
linelist <- rio::import(here::here("data", "linelist_cleaned.rds"))
```

Now type code that will print the first 10 lines of data in the `linelist` data frame.  

```{r exercise2-head, exercise = TRUE}
```

```{r exercise2-head-solution}
head(linelist)
```



## Video

Testing the embed of a video

Watch this video about data cleaning and management

<iframe width="560" height="315" src="https://www.youtube.com/embed/nxBTDfJ_neE" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>



## Topic 2

### Exercise with Hint

*Here's an exercise where the chunk is pre-evaulated via the `exercise.eval` option (so the user can see the default output we'd like them to customize). We also add a "hint" to the correct solution via the chunk immediate below labeled `print-limit-hint`.*

Modify the following code to limit the number of rows printed to 2:

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
head(linelist)
```

```{r print-limit-hint}
Specify the desired number of rows after a comma
```

```{r print-limit-solution}
head(linelist, 2)
```

### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```
